Scheduling and Vectorization for MPC

### 2022-03-22 overleaf/prob_stmt_np_hardness.tex
updated NP-Hardness proof, added "loud thinking" section

In order to benefit from parallelization/amortization, we must schedule two or more $A$-instructions in the same parallel node (or two or more $M$-instructions in the same parallel node). We also assume that scheduling $A$-instructions in parallel with $M$-instruction does not benefit from amortization\footnote{this is not strictly true, but assuming it, e.g. as in \cite{Ishaq2019, Demmler2015ABYA, Mohassel2018}, helps with the exposition.}. It incurs the exact same cost as scheduling the $A$-instructions in a node $P_A$, scheduling the $M$-instructions in a node $P_M$, and having $P_A$ precede $P_M$ in the parallel schedule. We use the following cost model:

@@

% Given a serial schedule (a linear graph) of an MPC program i.e. a sequence of instructions $\mathcal{S}:=(S_1, \dots, S_n)$, where $S_i \in \{A, M\}, 1 \leq i \leq n$, and a def-use dependency graph $G(V, E)$ corresponding to $\mathcal{S}$, our task is to construct a parallel schedule (another linear graph) $\mathcal{P} := (P_1, \dots, P_n)$ observing the following conditions:

% \begin{enumerate}
%     \item Multiple, not necessarily continuous instructions of the same type (i.e. either $A$ or $M$) from $\mathcal{S}$ can be grouped into a single $P_i$.
%     \item Def-use dependencies of the graph $G(V, E)$ should be preserved i.e. if instructions $S_i, S_j, i < j$ are a def-use (an edge exists from $S_i$ to $S_j$ in $G$), then they can only be mapped to $P_{i'}, P_{j'}, i' < j'$.
% \end{enumerate}

% Our goal is to construct minimize the height of the graph $\mathcal{P}$. Indeed, a graph $\mathcal{P}$ with minimum height will maximize parallelization.

\ishaq{Should be better to model this as a tree, and the branches (before they merge) can be executed in parallel, Right now it is too imprecise. Ideas please?}
Consider an MPC program as a set of $n$ sequences: $\{S_1, \dots, S_n\}$. Each $S_i$ is a tuple of $A$ and $M$ instructions.\footnote{Every MPC program can be transformed into such a set of sequences.} All elements of the set $\{S_1, \dots, S_n\}$ can execute in parallel, however, all instructions within an element (i.e. sequence) must execute sequentially. For example, consider the three sequences, right arrow indicates a \emph{dependence}, meaning that the source node must execute before the target node:

@@

The cost of a schedule $P$ is

\begin{equation}
    \mathit{cost}(P) = \sum_{i=1}^k \mathit{cost}(P_i)
\end{equation}

where $\mathit{cost}(P_i) = \alpha$ if $P_i$ consists of $A$-instructions only, it is $\beta$ if $P_i$ consists of $M$-instructions only, and it is $(\alpha + \beta)$ if $P_i$ mixes $A$-instructions and $M$-instructions.

@@

\paragraph{Correctness} Correctness of $\mathcal{P}$ is guaranteed by definition. Since \emph{dependencies} are preserved, the computed function remains the same.

% Note that we use an MPC-Source control flow graph (CFG) $G'(V', E')$ along with def-use graph $G(V,E)$ to construct $\mathcal{P}$. We consider a linearized MPC schedule $\mathcal{S}$ above for ease of exposition. The argument becomes slightly more involved when dealing with a graph $G'$ that may contain cycles.

\ishaq{TODO: rather than multiple amortized instructions costing the same as a single non-amortized instruction, Assume a function monotonically decreasing function for costs and prove the theorem using that function.}

The problem of finding a schedule $P$ with a minimal $cost(P)$ for a given loop body has been shown to be an NP-Hard problem, as it can be reduced to the problem of finding a \emph{shortest common supersequence}, a known NP-Hard problem\cite{Maier1978},\cite{Vazirani2010}. The shortest common supersequence problem is as follows: {\it given two or more sequences find the the shortest sequence that contains all of the original sequences.} This can be solved in $O(n^k)$ time, where $n$ is the cardinality of the longest sequence and $k$ is the number of sequences. For our problem $n$ is the maximum length of a node and $k$ is the number of total number of nodes.

To see the reduction, suppose $P$ is a schedule with minimal cost (computed by a black-box algorithm). We can derive a schedule $P'$ with the same cost as $P$, by mapping each mixed node $P_i \in P$ to two consecutive nodes in $P'$: an $A$-instruction node followed by an $M$-instruction node. Clearly, $P'$, which now is a sequence of $A$'s and $M$'s, is a supersequence of each sequence $S_i$, i.e., $P'$ is a common supersequence of $S_1 \dots S_n$. It is also a shortest common supersequence. To see this, let $X$ and $Y$ denote, respectively, the number of $A$ and $M$ nodes in $P'$. The cost of $P'$, and $P$, is $X \cdot \alpha + Y \cdot \beta$. Now suppose, there exists a shorter common supersequence, $P''$ that consists of $X'$ nodes of type $A$-instructions $Y'$ nodes of type $M$-instructions. Since $P''$ is shorter than $P'$, therefore $X' + Y' < X + Y$, and $X' \cdot \alpha + Y' \cdot \beta < X \cdot \alpha + Y \cdot \beta$ i.e. $\mathit{cost}(P'') < \mathit{cost}(P')$. But $\mathit{cost}(P') = \mathit{cost}(P)$ and $\mathit{cost}(P)$ is the optimal cost.  Therefore $\mathit{cost}(P'') < \mathit{cost}(P')$ is contradiction and no such $P''$ exists.

>>>>

Given a serial schedule (a linear graph) of an MPC program i.e. a sequence of instructions $S := (S_1, \dots, S_n)$, where $S_i \in \{A, M\}, 1 \leq i \leq n$, and a def-use dependency graph $G(V, E)$ corresponding to $S$, our task is to construct a parallel schedule (another linear graph) $P := (P_1, \dots, P_m)$ observing the following conditions:

\begin{enumerate}
    \item Multiple, not necessarily continuous instructions from $S$ can be grouped into a single $P_i$.
    \item Def-use dependencies of the graph $G(V, E)$ should be preserved i.e. if instructions $S_i, S_j, i < j$ form a def-use i.e. an edge exists from $S_i$ to $S_j$ in $G$), then they can only be mapped to $P_{i'}, P_{j'}$ such that $i' < j'$.
\end{enumerate}

\paragraph{Correctness} Correctness of $\mathcal{P}$ is guaranteed by definition. Preserving def-use \emph{dependencies} means the computed function remains the same in both $S$ and $P$.

In order to benefit from parallelization/amortization, we must schedule two or more $A$-instructions in the same parallel node (or two or more $M$-instructions in the same parallel node). We also assume that scheduling $A$-instructions in parallel with $M$-instruction does not benefit from amortization\footnote{this is not strictly true, but assuming it, e.g. as in \cite{Ishaq2019, Demmler2015ABYA, Mohassel2018}, helps simplify the problem.}. It incurs the exact same cost as scheduling the $A$-instructions in a node $P_A$, scheduling the $M$-instructions in a node $P_M$, and having $P_A$ precede $P_M$ in the parallel schedule. We use the following cost model:

@@

The cost of schedule $S$ is
\begin{equation}
    \mathit{cost}(S) = \sum_{i=1}^n \mathit{cost}(S_i)
\end{equation}

where $cost(S_i) = \alpha$ if $S_i$ is an $A$-instruction and $cost(S_i) = \beta$ if $S_i$ is an $M$-instruction. Similarly, the cost of schedule $P$ is

\begin{equation}\label{eq_par_schedule_cost}
    \mathit{cost}(P) = \sum_{i=1}^m \mathit{cost}(P_i)
\end{equation}

Each $P_i$ may contain multiple instructions, thus $\mathit{cost}(P_i) = \alpha$ if $P_i$ consists of $A$-instructions only, it is $\beta$ if $P_i$ consists of $M$-instructions only, and it is $(\alpha + \beta)$ if $P_i$ mixes $A$-instructions and $M$-instructions. Our goal is to construct a parallel schedule $P$ that reduces the program cost (when compared to cost of $S$).

Note that we consider a linearized MPC schedule $S$ above for ease of exposition only. In our tool-chain we use an MPC-Source control flow graph (CFG) $G'(V', E')$ along with def-use graph $G(V,E)$ to construct $P$. The argument becomes slightly more involved when dealing with a graph $G'$ that may contain cycles.

To prove that optimal scheduling is an NP-Hard problem, we consider the following convenient representation. An MPC program is represented as a set of sequences $S = \{S_1, \dots, S_n\}$. Each element $S_i \in S$ is a tuple. The items of the tuple $S_i$ are operations, i.e. $A$ or $M$ instructions, that have to be executed in order (operations depend on previous operations i.e. $S_i[j], j > 1$ depends $S_i[j-1]$). However, the sequences $S_i, 1 \leq i \leq n$ themselves, can overlap each other in any way i.e. distinct sequences can be executed in parallel. We argue that an MPC program can be transformed into such collection of sequences by traversing the circuit for each pair of input and output values. \ishaq{I am not sure how this will be done, see \cref{fig:sequence_construction}}

As an example, consider the MPC program consisting of the following three sequences, right arrow indicates a \emph{dependence}, meaning that the source node must execute before the target node:

@@

Cost of schedule $P$ is computed using \cref{eq_par_schedule_cost} above.
% The cost of a schedule $P$ is

% \begin{equation}
%     \mathit{cost}(P) = \sum_{i=1}^k \mathit{cost}(P_i)
% \end{equation}

% where $\mathit{cost}(P_i) = \alpha$ if $P_i$ consists of $A$-instructions only, it is $\beta$ if $P_i$ consists of $M$-instructions only, and it is $(\alpha + \beta)$ if $P_i$ mixes $A$-instructions and $M$-instructions.

@@

The problem of finding a schedule $P$ with a minimal $cost(P)$ for a given loop body has been shown to be an NP-Hard problem, as it can be reduced to the problem of finding a \emph{shortest common supersequence}, a known NP-Hard problem\cite{Maier1978, Vazirani2010}. The shortest common supersequence problem is as follows: {\it given two or more sequences find the the shortest sequence that contains all of the original sequences.} This can be solved in $O(n^k)$ time, where $n$ is the cardinality of the longest sequence and $k$ is the number of sequences. For our problem $n$ is the maximum length of a node and $k$ is the number of total number of nodes.

To see the reduction, suppose $P$ is a schedule with minimal cost (computed by a black-box algorithm). We can derive a schedule $P'$ with the same cost as $P$, by mapping each mixed node $P_i \in P$ to two consecutive nodes in $P'$: an $A$-instruction node followed by an $M$-instruction node. Clearly, $P'$, which now is a sequence of $A$'s and $M$'s, is a supersequence of each sequence $S_i$, i.e., $P'$ is a common supersequence of $S_1 \dots S_n$. It is also a shortest common supersequence. To see this, let $X$ and $Y$ denote, respectively, the number of $A$ and $M$ nodes in $P'$. The cost of $P'$, and $P$, is $X \cdot \alpha + Y \cdot \beta$. Now suppose, there exists a shorter common supersequence, $P''$ that consists of $X'$ nodes of type $A$-instructions $Y'$ nodes of type $M$-instructions. Since $P''$ is shorter than $P'$, therefore $X' + Y' < X + Y$, and $X' \cdot \alpha + Y' \cdot \beta < X \cdot \alpha + Y \cdot \beta$ i.e. $\mathit{cost}(P'') < \mathit{cost}(P')$. But $\mathit{cost}(P') = \mathit{cost}(P)$ and $\mathit{cost}(P)$ is the optimal cost.  Therefore $\mathit{cost}(P'') < \mathit{cost}(P')$ is contradiction and no such $P''$ exists. \qed

\subsection{Loud Thinking}
\ishaq{I will delete this section after we have decided on how to handle various issues raised in this section}.

\begin{figure}
  \includegraphics[width=0.6\linewidth]{overleaf/figs_paper_SIMD/sequence_construction.pdf}
  \caption{How do we construct sequences from this circuit?} 
  \label{fig:sequence_construction}
\end{figure}

The more realistic cost model is as under:
\begin{enumerate}
    \item The cost of an $A$-instruction is given by a function $\alpha(\cdot)$ where the only parameter to the function is the number of $A$ instructions that will be executed in parallel.
    \item Similarly, the cost of an $M$-instruction is given by function $\beta(\cdot)$.
    \item Both $\alpha$ and $\beta$ are amortization function e.g. $\alpha(n) \leq \alpha(\ell) + \alpha(m); n = \ell + m$. Same condition applies for $\beta$.
\end{enumerate}

\begin{figure}
  \includegraphics[width=0.6\linewidth]{overleaf/figs_paper_SIMD/cost_inequality.pdf}
  \caption{It is unclear whether right super sequence cost less/more/the-same as left one.} 
  \label{fig:cost_comparison}
\end{figure}

I am not using this cost model (for now) because NP-Hardness proof is tricky (see \cref{fig:cost_comparison}). Consider a super sequence that has nodes (of only one type) with following weights: $SS_1 = (1, 1, 4)$ where $Length(SS_1) = 3$ and $cost(SS_1) = \alpha(1) + \alpha(1) + \alpha(4)$.

Using $\alpha(4) \leq \alpha(1) + \alpha(3)$, we can say

\begin{equation}\label{ineq_cost_1}
    \alpha(1) + \alpha(1) + \alpha(4) \leq \alpha(1) + \alpha(1) + \alpha(1) + \alpha(3)
\end{equation}

Now suppose, for the same schedule, another super sequence with weights $SS_2 = (3, 3)$ exists, $Length(SS_2) = 2$ and $cost(SS_2) = \alpha(3) + \alpha(3)$. Using $\alpha(3) \leq \alpha(1) + \alpha(1) + \alpha(1)$, we can say

\begin{equation}\label{ineq_cost_2}
    \alpha(3) + \alpha(3) \leq \alpha(1) + \alpha(1) + \alpha(1) + \alpha(3)
\end{equation}

The problem is that while RHS is the same in \cref{ineq_cost_1} and \cref{ineq_cost_2}, we cannot say anything about the relationship between their LHS. It could be possible that cost of $SS_2$, LHS of  \cref{ineq_cost_2}, is more than cost of $SS_1$, LHS of \cref{ineq_cost_1}. Thus, we have a shorter super sequence $SS_2$ that does not contradict that the optimality of the schedule used to generate $SS_1$.
